---
title: "GroupProject"
author: "Casey Hicks"
date: "2023-11-05"
output:
  html_document: 
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

ADW = Average of aggravated assaults 2005/2006
MPD = Metro police defined disorder: MPD Calls for Service "Disorderly"
UIdis = MPD Calls for Service: "MAN DOWN" "WOMAN DOWN" "SHOOTING" "SOUNDS OF GUNSHOTS" "INDECENT EXPOSURE" "SOLICIT PROSTITUTION / SLIP" "SOLICITING FOR PROSTITUTION" "DESTRUCTION OF PROPERTY"
DV = Average of domestic violence calls for service


```{r fig1, fig.height = 10, fig.width = 12, fig.align = "center"}

library(dplyr)

##### FUNCTIONS ######

# Group data by TRACT and rename rownames by that column
# Parameters: group_data(data-to_group, new_df_name)
group_data <- function(data,new_df_name) {
  sorted <- data %>%
    group_by(TRACT) %>%
    summarise(across(where(is.numeric), sum), .groups = "drop")
  df <- as.data.frame(sorted)
  rownames(df) <- df[, 1]
  df <- df[, -1]
  assign(new_df_name, df, envir = .GlobalEnv)
}

# Remove outliers
remove_outliers <- function(data, outliers) {
  data <- data[!(rownames(data) %in% outliers), ]
}

######################

#Read the data:
orig_data <- read.csv("/Users/casey.hicks/Desktop/Multivariate Analysis/Project/Data/AlcoholData.csv")

# removed location ID numbers
data <- subset(orig_data, select = -c(STATE_FIPS,TRACT,OBJECTID,BLKGRP))


######################

# Data2: All chosen group variables
data2 <- orig_data[,c("TRACT", "VACANTPERCENT","POP2004","POP04_SQMI","MALE","YOUNGPOP","PROSOCIAL",
                      "METRO_BG","STORE_BG","NIGHT_BG","TAVERN_BG","REST_BG",
                      "PHYSDIS0506","AVGDV","AVGMPDDIS","AVGUIDISOR","AVGADW0506","ARR_0506")]

group_data(data2, "data2")

outliers <- c("5800", "4400", "5402", "5500", "202", "3800", "100")
#remove_outliers(data2, outliers) NOT WORKING??

data2 <- data2[!(rownames(data2) %in% c("5800", "4400", "5402", "5500", "202", "3800", "100")), ]


data2.pca <- princomp(data2, cor=TRUE)               
summary(data2.pca, loadings=TRUE) 

colors_pca <- sapply(rownames(data2.pca$scores), function(rowcolors) {
  rownum <- as.numeric(rowcolors)
  if (rownum >= 0 && rownum <= 999) {
    return("pink")
  } else if (rownum >= 1000 && rownum <= 1999) {
    return("violet")
  } else if (rownum >= 2000 && rownum <= 2999) {
    return("purple")
  } else if (rownum >= 3000 && rownum <= 3999) {
    return("blue")
  } else if (rownum >= 4000 && rownum <= 4999) {
    return("cyan")
  } else if (rownum >= 5000 && rownum <= 5999) {
    return("green")
  } else if (rownum >= 6000 && rownum <= 6999) {
    return("yellow")
  } else if (rownum >= 7000 && rownum <= 7999) {
    return("orange")
  } else if (rownum >= 8000 && rownum <= 8999) {
    return("red")
  } else if (rownum >= 9000 && rownum <= 9999) {
    return("brown")
  } else {
    return("gray")
  }
})

biplot(data2.pca, col = c("black", "black"), cex = 0.5, xlim = c(-0.15, 0.3))
text(
  data2.pca$scores,           
  labels = rownames(data2.pca$scores), 
  col = colors_pca,
  cex = 0.6,                           
  pos = 4                              
)



colors <- sapply(rownames(data2), function(rowcolors) {
  rownum <- as.numeric(rowcolors)
  if (rownum >= 0 && rownum <= 999) {
    return("pink")
  } else if (rownum >= 1000 && rownum <= 1999) {
    return("violet")
  } else if (rownum >= 2000 && rownum <= 2999) {
    return("purple")
  } else if (rownum >= 3000 && rownum <= 3999) {
    return("blue")
  } else if (rownum >= 4000 && rownum <= 4999) {
    return("cyan")
  } else if (rownum >= 5000 && rownum <= 5999) {
    return("green")
  } else if (rownum >= 6000 && rownum <= 6999) {
    return("yellow")
  } else if (rownum >= 7000 && rownum <= 7999) {
    return("orange")
  } else if (rownum >= 8000 && rownum <= 8999) {
    return("red")
  } else if (rownum >= 9000 && rownum <= 9999) {
    return("brown")
  } else {
    return("gray")
  }
})



data2.d <- dist(scale(data2), diag=TRUE, upper=TRUE)
cmd2 <- cmdscale(data2.d, eig=TRUE)


plot(cmd2$points[,1:2], pch = "", main = "MDS", xlab = "MDS Points: 1", ylab = "MDS Points: 2")
text(cmd2$points[,1:2], labels = rownames(data2), col = colors, cex = 0.6)

# 
# data2.pca$loadings
# 
# 
# data3 <- orig_data[,c("TRACT","STORE_BG","NIGHT_BG","TAVERN_BG","REST_BG","AVGDV","AVGMPDDIS","AVGUIDISOR","AVGADW0506")]
# head(data3) 
# 
# cor <- cor(data3)
# 
# 
# d3 <- data3 %>%
#    group_by(TRACT) %>%
#    summarise(across(where(is.numeric), sum), .groups = "drop")
# 
# #data3 <- na.omit(data3)
# data3 <- as.matrix(data3)
# rownames(data3) <- data3[,1]
# data3 <- (data3[,-1])
# 
# data3["100",]
# 
# data3.pca <- princomp(data3, cor=TRUE)
# summary(data3.pca, loadings = TRUE)
# 
# biplot(data3.pca, col = c("black", "red"), cex =0.7)
# 
# data3.dist <- dist(scale(data3), diag = TRUE, upper = TRUE)
# 
# data3.mds <- cmdscale(data3.dist, eig=TRUE)
# #data3.mds
# 
# plot(data3.mds$points, type = "n", main="MDS Analysis (Correlation Distance)", xlab="1st Component", ylab="2nd Component")
# text(data3.mds$points, labels=rownames(data3), cex=0.7)
# 



```
<br>

#### PCA of crimes only
```{r}
data4 <- subset(orig_data, select =
                  c(TRACT,PHYSDIS0506,AVGDV,AVGMPDDIS,AVGUIDISOR,AVGADW0506,ARR_0506))

d4 <- data4 %>%
   group_by(TRACT) %>%
   summarise(across(where(is.numeric), sum), .groups = "drop")

data4 <- as.matrix(d4)
rownames(data4) <- data4[,1]
data4 <- (data4[,-1])

data4.pca <- princomp(data4, cor=TRUE)               
summary(data4.pca, loadings=TRUE)               
biplot(data4.pca, col = c("black", "red"), cex =0.7)


# Cor matrix for realationship between types of crime
# High cor between DV, UI, and ADW
# Lower cor between physdis and violent crimes 
cor(data4)

```

#### PCA crimes and alcohol establishments
```{r fig, fig.height = 10, fig.width = 10, fig.align = "center"}
data5 <- subset(orig_data, select =
                  c(TRACT,PHYSDIS0506,AVGDV,AVGMPDDIS,AVGUIDISOR,AVGADW0506,ARR_0506,
                    ON_SQMI,OFF_SQMI,REST_SQMI,TAVERN_SQMI,NIGHT_SQMI,STORE_SQMI))

cor(data5)

d5 <- data5 %>%
   group_by(TRACT) %>%
   summarise(across(where(is.numeric), sum), .groups = "drop")

data5 <- as.data.frame(d5)
rownames(data5) <- data5[,1]
data5 <- (data5[,-1])

outliers5 <- c("4002","5500", "5800", "2101")
data5 <- data5[!(rownames(data5) %in% outliers5), ]
### data5 is complete


data5.pca <- princomp(data5, cor=TRUE)               
summary(data5.pca, loadings=TRUE)               
biplot(data5.pca, col = c("black", "red"), cex =0.7, xlim = c(-0.1,0.3))

data5.d <- dist(scale(data5), diag=TRUE, upper=TRUE)
cmd5 <- cmdscale(data5.d, eig=TRUE)
#cumsum(cmd5$eig)/sum(cmd5$eig)


#YOUNGPOP color filter
d5b <- orig_data %>%
   group_by(TRACT) %>%
   summarise(across(where(is.numeric), sum), .groups = "drop")

data5b <- as.data.frame(d5b)
rownames(data5b) <- data5b[,"TRACT"]
data5b <- (data5b[,-1])
data5b <- data5b[!(rownames(data5b) %in% outliers5), ]

###





median5 <- median(orig_data$YOUNGPOP)
colors5 <- ifelse(data5b$YOUNGPOP < median5, "blue", "red")

plot(cmd5$points[,1:2], pch = "", main = "MDS crime vs alcohol estab.")
text(cmd5$points[,1:2], labels = rownames(data5), col = colors5, cex = 0.6)

data5.cor <- cor(data5)
data5.cordist <- 1-data5.cor

cmd5.cor <- cmdscale(data5.cordist, k = 2, eig = TRUE)
plot(cmd5.cor$points[,1:2], pch = "", main = "MDS (Corr Dist) crime vs alcohol estab.")
text(cmd5.cor$points[,1:2], labels = rownames(data5.cordist), cex = 0.6)

```

### K-Means Clustering
```{r}
#install.packages("tidyverse")
#install.packages("factoextra")

library(tidyverse)  # data manipulation
library(cluster)    # clustering algorithms
library(factoextra) # clustering algorithms & visualization


k2 <- kmeans(data5, centers = 8, nstart = 25)
fviz_cluster(k2, data = data5)

#k2
#aggregate(data2, by=list(cluster=k2$cluster), mean)



```
